TO FIND

### 2013-05-27 ms.tex
Update ms.tex

\textbf{Title:}   Using neutral theory to reveal the contribution of dispersal to community assembly in complex landscales\\

@@

Ecologists are required to move toward a predictive ecology, integrating
elements of theoretical ecology \parencite{Thuiller2013}, and the metacommunity
perspective appears naturally as the appropriate conceptual framework to develop
the new modeling techniques required to fill this challenge. The metacommunity
concept builds on feedbacks between local scale processes, such as competitive
interactions and local adaptation, and regional scale processes such as
dispersal, gene flow and speciation \textcite{Leibold2004a}. This perspective is
particularly relevant to limnology, where exchanges of organisms and nutrients
affect community and ecosystem properties from the local (e.g. vertical mixing
\parencite{Ryabov2011}) to the regional (e.g. connection of lakes
\parencite{Leibold2004b} scales. It emphasizes the importance of dispersal
relative to pairwise interactions in the organization of ecological communities.

The development of neutral theory has been quite provocative in that respect, as
one could see it as a step back in time. Neutral theory makes the provocative
assumption that species are ecologically equivalent and thereby any variation in
the environment has no impact on demography \parencite{Bell2000,Hubbell2001}.
Only demographic stochasticity and dispersal drive the structure of neutral
ecological communities. It therefore appears that, on first sight, neutral
theory is useless.  We will develop in this paper the argument that neutral
theory could be a useful tool to understand the impact of dispersal on community
organization in landscapes of various complexities.

Neutral theory sparked an historical debate that is still lasting after more
than a decade \parencite{Chave2004, Etiennee2011, Rosindell2012,Clark2012}. It
was stimulated by the surprising ability of neutral models to fit several well
studied empirical observations such as species abundance distributions and
distance-decay relationships. A remarkable strenght of neutral theory is to
provide a \emph{"formal general theory of abundance and diversity that will
account, in a simple and economical fashion, for the many patterns that
ecologists have documented"} \parencite{Bell2001}. Even if new studies rejecting
neutral theory are consistently published (e.g.  \textcite{Ricklefs2012}), there
is now almost a consensus that neutral theory is a well-developed null
hypothesis for niche theory and could even be used as an adequate approximation
of ecological dynamics in some situations.  \textcite{Bell2001} nicely
envisionned two perspectives to neutral theory that are still standing today.
Under the weak perspective, neutral theory provides a set of realistic
predictions of community organization despite false assumptions. Even if being
fundamentally wrong, neutral theory is still useful when used as a null
hypothesis \parencite{Gotelli2006}. It is considered as an improvement over
traditional null hypotheses based on randomization \parencite{Gotelli2000}
because it readily integrates dispersal. The strong version on the other hand
posits that neutral theory is a satisfying approximation to community dynamics
and an appropriate theory to explain the distribution of biodiversity. It
implies that the right mechanisms have been identified and that the consistently
observed differences among species do not impact community organization.

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2011}. Because every
ecological model is a simplification of reality, any scientist has to
subjectively decide the level of details he puts in, leaving out some elements
judged unimportant. The realism perspective requires that all assumptions of
theory to be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or prediction (another old philosophical debate, see
\textcite{Schmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2011} is a satisfying approximation.

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? We see two
potential answers to this question. A first answer might be that stochasticity
of various origins can blur the deterministic differences among species and
promote ecological drift \parencite{Gravel2011}. Much has been said the
existence of demographic stochasticity, some ecologists even arguing that
neutral models impede progress in community ecology by hidding niche differences
\parencite{Clark2012}, and we therefore will let this discussion for other
papers. The second answer is that dispersal and historical contingencies might
have a much more profound impact on species distribution \parencite{Bahn2007,
Boulangeat2012} and ecological dynamics. The debate over the equivalence
assumption and demographic stochasticity has perhaps overlook the recognition of
how much dispersal influence community assembly.

In this paper, we argue that neutral theory is useful both to understand and
predict the impact of dispersal on community organization. Even for purely
theoretical analyses, we need a benchmark without niche differences to reveal
the role of dispersal in structuring communities, and understand how it
interacts with niche differentiation. We will explore recent applications of
neutral theory, at the crossroad of network theory, to better represent the
impact of landscape structure on biodiversitity distribution. This analysis will
prove particularly relevant to limnology, where most riverine and lacustre
habitats are characterized by a their discrete nature and spatially complex
arrangements \parencite{Peterson2013}. We will also reveal the relative
contribution of ecological interaction and niche differentiation by contrasting
predictions of a neutral model to other metacommunity perspectives.

Our main objective in this paper is to use neutral theory to stress the
importance of landscape network structure on the distribution of diversity. We
refer to the landscape organization as a "spatial contingency"
\parencite{Peres-Neto2013} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or spatially explicity (e.g. over a lattice), and spatial
constant, to a perspective focusing on the variance of dispersal. A second
generation of neutral models (e.g. \textcite{Economo2008,
Economo2011,Desjardins2012a,Desjardins2012b}, and even experiments
\parencite{Altermat2012}, recently introduced more realistic lanscapes and found
surprising contributions of spatial contingencies. We will start with a short
review of the main approaches to describe spatial networks and the studies
investigating them. Then we will describe three simple toy models of
metacommunity dynamics, taking this opportunity to review their assumptions and
main predictions. We provide as Supplementary Material the R scripts for the toy
models and all simulations conducted for this paper. We then conduct simple
simulations of these models to reveal with simple examples the impact of spatial
network structure on diversity distribution. We conclude with a discussion on
the operationally of the framework.

@@

A network is a discrete mathematical object made of two sets: a set of
vertices (or nodes) $V$ and a set of edges $E$ connecting the vertices
\cite{new10}. The term ``graph'' is often preferred in computer science and
mathematics \cite{gro06}, with graph algorithms being an important and active
area of research \cite{sed01}. A network is a combinatorial object: it is
used to study how discrete entities are connected and how they combine
together to create complex structures. They are used to study molecules, food
webs, social networks, or even the relationship between variables in
statistics \cite{wri21,new10}. We are especially interested in spatial
networks, a special kind of network mixing the combinatorial properties of
networks with a topological space \cite{kob94}. Thus, the vertices in a
spatial graph are embedded in some other space, most often the two or three-
dimensional Euclidean space. This object brings a rich representation to
spatial ecology and is particularly suited for systems of lakes and rivers, which can
easily be represented by vertices and edges. There are two notions of
distance in spatial networks. Euclidean distance represents the geographical
distance between the vertices $(i, j)$, i.e.: $\sqrt{(x_i - x_j)^2 + (y_i - y_
j)^2}$. Geodesic distance is the distance in the graph space, i.e.: the
length of the shortest path \cite{dij59}. For example, two lakes could be
very close on a map (short Euclidean distance) but if they are not directly
linked by a river the geodesic distance could be great.

@@

We also provide the code for a second structure that we call a random geometric
tree.  It builds a tree from the shortest path tree \cite{dij59} of a vertex in
a random geometric network. This random geometric tree does not exaxctly
represent dendritic landscapes but is a convenient model to simulate lake
connected by rivers to a series of smaller lakes.

@@

% TO PHIL: IT WOULD BE GREAT IF YOU COULD EXPAND ONE OR TWO PARAGRAPHS ON THE 
% MAIN STATISTICS USED TO DESCRIBE SPATIAL GRAPHS. AT LEAST PRESENT THE MAIN 
% DIFFERENCE BETWEEN NETWORK AND VERTEX LEVEL METRICS, AND THEN REFER TO THE 
% TABLE.

>>>>

\textbf{Title:}   Using neutral theory to reveal the contribution of dispersal to community assembly in complex landscapes\\

@@

Ecology needs to move toward a more predictive approach. Integrating
elements of theoretical ecology \parencite{Thuiller2013} with the metacommunity
perspective appears naturally as a good candidate to fill this challenge. The metacommunity
concept builds on feedbacks between local scale processes, such as competitive
interactions and local adaptation, and regional scale processes such as
dispersal, gene flow and speciation \textcite{Leibold2004a}. This perspective is
particularly relevant to limnology, where exchanges of organisms and nutrients
affect community and ecosystem properties from the local (e.g. vertical mixing
\parencite{Ryabov2011}) to the regional (e.g. connection of lakes
\parencite{Leibold2004b} scales. It emphasizes the importance of dispersal
relative to pairwise interactions in the organization of ecological communities.

At first sight, the development of neutral theory appears as a step back. Based on Kimura's neutral theory of molecular evolution, the neutral theory of biodiversity makes the provocative
assumption that species are ecologically equivalent. Thus, variation in
the environment has no impact on demography \parencite{Bell2000,Hubbell2001}. Neutral
ecological communities are driven only by demographic stochasticity and dispersal. Neutral theory sparked an historical debate still lasting after more
than a decade \parencite{Chave2004, Etiennee2011, Rosindell2012,Clark2012}. It
was stimulated by the surprising ability of neutral models to fit some well
studied empirical observations such as species abundance distributions and
distance-decay relationships. A remarkable strength of the theory is to
provide a \emph{"formal general theory of abundance and diversity that will
account, in a simple and economical fashion, for the many patterns that
ecologists have documented"} \parencite{Bell2001}. Even if new studies rejecting
neutral theory are consistently published (e.g.  \textcite{Ricklefs2012}), a consensus is forming that neutral theory is a well-developed null
hypothesis for niche theory and could even be used as an adequate approximation
of ecological dynamics in some situations.  \textcite{Bell2001} nicely
envisioned two perspectives to neutral theory that are still standing today.
Under the weak perspective, neutral theory provides a set of realistic
predictions of community organization despite false assumptions. Even if being
fundamentally wrong, neutral theory is still useful when used as a null
hypothesis \parencite{Gotelli2006}. It is considered as an improvement over
traditional null hypotheses based on randomization \parencite{Gotelli2000}
because it readily integrates dispersal. The strong version on the other hand
posits that neutral theory is a satisfying approximation to community dynamics
and an appropriate theory to explain the distribution of biodiversity. It
implies that the right mechanisms have been identified and that the consistently
observed differences among species do not impact community organization.

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2011}. Because every
ecological model is a simplification of reality, scientists have to
subjectively decide the level of details they put in, leaving out some elements they consider unimportant. The realism perspective requires that all assumptions of
theory must be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or prediction (another old philosophical debate, see
\textcite{Schmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2011} is a satisfying approximation.

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? Perhaps stochasticity
of various origins blur the deterministic differences among species and
promote ecological drift \parencite{Gravel2011}. A second answer is that dispersal and historical contingencies might
have a much more profound impact on species distribution \parencite{Bahn2007,
Boulangeat2012} and ecological dynamics. The debate over the equivalence
assumption and demographic stochasticity has perhaps overlook the recognition of
how much dispersal influence community assembly.

In this paper, we argue that neutral
theory can be a useful tool to understand the impact of dispersal on community
organization in landscapes of various complexities. Even for purely
theoretical analyses, we need a benchmark without niche differences to reveal
the role of dispersal in structuring communities and understand how it
interacts with niche differentiation. We will explore recent applications of
neutral theory, at the crossroad of network theory, to better represent the
impact of landscape structure on biodiversity distribution. This analysis will
prove particularly relevant to limnology, where most riverine and lacustre
habitats are characterized by their discrete nature and spatially complex
arrangements \parencite{Peterson2013}. We will also explore the relative
contribution of ecological interaction and niche differentiation by contrasting
predictions of a neutral model to other metacommunity perspectives.

Our main objective is to use neutral theory to stress the
importance of landscape network structure on the distribution of diversity. We
refer to the landscape organization as a "spatial contingency"
\parencite{Peres-Neto2013} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or spatially explicit (e.g. over a lattice), and spatial
constant, to a perspective focusing on the variance of dispersal. A second
generation of neutral models (e.g. \textcite{Economo2008,
Economo2011,Desjardins2012a,Desjardins2012b}, and even experiments
\parencite{Altermat2012}, recently introduced more realistic landscapes and found
surprising contributions of spatial contingencies. We start with a short
review of the main approaches to describe spatial networks. Then we describe three simple toy models of
metacommunity dynamics, using this opportunity to review their assumptions and
main predictions. We provide as Supplementary Material the R scripts for the toy
models and all simulations conducted for this paper. We then conduct simple
simulations of these models to reveal with simple examples the impact of spatial
network structure on diversity distribution. We conclude with a discussion on
the operationally of the framework.

@@

A network is a discrete mathematical object made of two sets: a set of
vertices (or nodes) $V$ and a set of edges $E$ connecting the vertices
\cite{new10}. The term ``graph'' is often preferred in computer science and
mathematics \cite{gro06}, with graph algorithms being an important and active
area of research \cite{sed01}. A network is a combinatorial object: it is
used to study how discrete entities are connected and how they combine
together to create complex structures. They are used to study molecules, food
webs, social networks, or even the relationship between variables in
statistics \cite{wri21,new10}. We are especially interested in spatial
networks, a special kind of network mixing the combinatorial properties of
networks with a topological space \cite{kob94}. Thus, the vertices in a
spatial graph are embedded in some other space, most often the two or three-
dimensional Euclidean space. This object brings a rich representation to
spatial ecology and is particularly suited for systems of lakes and rivers, which can
easily be represented by vertices and edges. There are two notions of
distance in spatial networks. Euclidean distance represents the geographical
distance between the vertices $(i, j)$, i.e.: $\sqrt{(x_i - x_j)^2 + (y_i - y_
j)^2}$. Geodesic distance is the distance in the graph space, i.e.: the
length of the shortest path \cite{dij59}. For example, two lakes could be
very close on a map (short Euclidean distance) but the geodesic distance could be great if they are not directly
linked by a river.

The popularity of network theory stem for their ability to model complex
structures while allowing us to extract useful metrics (Table 1). At a very high level, 
a network can be described by a number of vertices (the order) and a number
of edges (the size). Looking more closely, the relationship between vertices
is influenced by paths, which are ordered series of nodes.
Centrality is a central (no pun intended) concept in network theory, where it
can be seen to as a measure of ``importance''. The simplest measure
of the centrality of a node is its degree, which is the number of nodes directly connected to it. Of course, this is a very rough description of centrality. For
example, two lakes can have the same degree, with one being connected to
a small isolated cluster, while the other one is part of one of the biggest 
network of lake. In this case, measures of centrality like eigen-centrality will
weight the importance of the connection so a node connected to well-connected
nodes will have higher centrality than a node connected to isolated nodes.

@@

We also provide the code for a second structure that we call a random geometric
tree.  It builds a tree from the shortest path tree \cite{dij59} of a vertex in
a random geometric network. This random geometric tree does not exactly
represent dendritic landscapes but is a convenient model to simulate lake
connected by rivers to a series of smaller lakes.

@@
